---
title: "SNA R2: Fundamental Network Concepts"
subtitle: "No R in this script, just concepts"
author: "J. Alberto Espinosa"
date: "11/14/2025"

format: 
   epub:
     toc: true
     toc-depth: 2
     toc-expand: 2
     toc-location: right

fig-width: 10
fig-height: 6
fontsize: 10pt
code-overflow: wrap
geometry: "left = 2cm, right = 2cm, top = 2cm, bottom = 2cm"
echo: true
include: true
warning: false
message: false

editor: 
  markdown: 
    wrap: 70

abstract: This Quarto file contains an introduction to fundamental network and SNA concepts. I expand on all these concepts and discuss them in more detail in later scripts. All SNA4M R Quarto scirpts can be found at [https://github.com/jibe4fun/sna4m/tree/R-Code](https://github.com/jibe4fun/sna4m/tree/R-Code). You can also query all the lectures, R scripts and other material for this course on the [SNA4M ChatBot] (https://poe.com/SNA4M). If you are interested in predictive analytics and machine learning R code, please refer to my PAML4M book R companion in my GitHub site [https://github.com/jibe4fun/paml4m/tree/R-Code](https://github.com/jibe4fun/paml4m/tree/R-Code). If you are interested in Predictive Analytics and Machine Learning for Managers, please refer to my PAML4M book [Predictive Analytics and Machine Learning for Managers](https://www.amazon.com/dp/B0C2RX8NY1) or chat with my chatbot associated with this book at [PAML4M ChatBot](https://poe.com/PAML4M). 
---

**Notes:**

1.  I include some R code in this script to render outputs or graphs
    to complement the conceptual explanations provided in this script.
    All scripts with the comment \`# Illustration\` are meant to
    display graphs and outputs to complement the text narratives, and
    there is no need to pay attention to the code, unless you want to.

2.  Also, sometimes I use the parameter \`#\| eval: false\` inside
    some R code sections marked with \`{r}\`, which simply prevents
    the script in that chunk from executing when you render the script
    into a document, which I use to avoid cluttering the output with
    unnecessary displays, help information, notifications, etc. To
    explore these commands, simply go to the respective script line
    and run it with \`Ctrl-Enter\`.

# Network Representations

A **network** is a collection of **actors** and their respective
**ties**. The graph terms for actors and ties are **vertices** and
**edges**. Vertices are sometimes referred to as **nodes**. Any two
vertices form a **dyad**, which is the basic unit of analysis in SNA.
One excellent property of SNA and graph analytics methods is that the
data can be represented and analyzed visually or quantitatively.

A **visual** representation of graphs is called a **sociogram** in SNA
and a **graph** in graph theory.

```{r}
# Illustration

library(statnet) # See package installation instructions below  

soc.mat <- rbind(c(0, 1, 0, 1, 0, 0),
                 c(1, 0, 0, 1, 1, 0),
                 c(0, 0, 0, 1, 0, 0),
                 c(1, 1, 1, 0, 0, 0),
                 c(0, 1, 0, 0, 0, 1),
                 c(0, 0, 0, 0, 1, 0))  

gplot(soc.mat, displaylabels = T, gmode = "graph")
```

A graph can also be represented as a **quantitative** matrix called a
**sociomatrix** (also called an **adjacency matrix**), which makes the
power of matrix algebra available to graph analytics and SNA. Cell
values contain a **1** if the two related vertices have a tie (for
**binary** networks), a **value** representing the weight or strength
of the tie (for **weighted** networks), or **0** if the vertices don't
have a tie.

```{r}
soc.mat
```

**Diagonal values** are often omitted when vertices don't have
relationships of interest with themselves (i.e., recursive
relationships).

```{r}
soc.mat.nodiag <- soc.mat # To avoid altering the original matrix
diag(soc.mat.nodiag) <- "" # Replace all diagonals with blanks
print(soc.mat.nodiag, quote = F)
```

# Vertex Attributes

Vertices can have **attributes.** Some **vertex** attributes are
**collected** and some are **measured**:

-   **Quantitative:** e.g., age, income, education, etc.

-   **Categorical:** e.g., gender, profession, nationality, etc.

Some **vertex** attributes are **computed** with network metrics:

-   **Quantitative:** e.g., degree centrality, betweenness centrality,
    etc.

-   **Categorical** or **Membership:** e.g., isolation, clique
    membership, cluster, community, etc.

It is this ability to compute vertex metrics, in addition to the
measured attributes that provides the **enhanced analytical power** of
SNA methods.

# Edge Attributes

Edges can also have **attributes.** Some **edge** attributes are
**collected** and some are **measured**:

-   **Quantitative:** e.g., time working together with a colleague,
    email exchange volume with a friend, number of courses taken
    together with a peer, etc.

-   **Binary:** e.g., friends with a colleague ( yes/no), LinkedIn
    contact with someone (yes/no), X follower (yes/no) or follwed by
    (yes/no), etc.

Some edge attributes are **computed** from network metrics:

-   **Quantitative:** e.g., distance to another member, paths to a
    colleague, walks between two members, etc.

-   **Binary:** e.g., is a tie to a member reciprocal, are two
    colleagues co-members of a clique, group or community, etc.

It is this ability to collect, measure, and compute edge metrics that
provides the **enhanced analytical power** of SNA methods.

# Other Attributes

In addition to vertex and edge attributes, networks possess many other
attributes at various levels, such as:

-   **Sub-Group Level:** Ego/alter network membership, clique
    co-membership, cluster grouping, community membership, etc.

-   **Network Level:** Structure, hierarchy, centralization,
    isolation, size, number of components, etc.

# Types of Networks

There are multiple types of networks based on several characteristics,
such as:

-   **Topology:** that is, the configuration of the network, such as
    star, ring, hierarchical, fully connected, etc.

-   **Binary** (Yes/No, 0/1) vs. **Weighted** (i.e., edges have
    values):

    -   **Binary** networks are best for visualization and for
        computing node metrics, such as centralities. A binary network
        contains ties between members that have a relationship (e.g.,
        are friends) and no tie if the actors don't have a
        relationship (e.g., not friends).

    -   **Weighted** or **valued** networks are best for quantitative
        analysis (e.g., clustering).

    -   **Dichotomized** networks are those extracted from a weighted
        network. We often need to convert a weighted network into a
        binary network. We do that by creating a binary network
        containing a tie when the weight exceeds a certain
        **threshold** (i.e., **cut**) of interest.

-   **Undirected** vs. **Directed**:

    -   An **undirected** network is one in which the relationships
        does not have a direction. For example, how long have 2 people
        known each other has no direction because the value is the
        same for either person, consequently, the ties don't have
        arrowheads. Undirected networks yield **symmetrical**
        sociomatrices in which the values above the diagonal are the
        same as those below the diagonal.

    -   **A directed** network is one in which the direction of the
        relationship matters. For example, how much person A likes
        person B is not necessarily the same as how much person B
        likes person A. In this case, how much A likes B is
        represented as a tie with an arrow going from A to B. Directed
        networks yield **asymmetrical** sociomatrices in which the
        values above the diagonal are not necessarily the same as
        those below the diagonal. The relationship in an asymmetric
        sociomatrix is specified from row to column (e.g., how much
        the row member likes the column member).

-   **Simplex** vs. **Multiplex**: A simplex network is one
    representing a single relationship (e.g., friendship). A multiplex
    network contains two or more layers, one for each relationship
    represented (e.g., friendship, advice, supervision, etc.).

-   **1-Mode** vs. **2-Mode**: In a **1-mode** network, the row actors
    are the same as the column actors (e.g., a group of classmates),
    and the cells represent the relationship between row and column
    actors. In a 2-mode network the row and column actors are
    different (e.g., rows are students and the columns are courses
    taken by students). A **2-mode** network can be be used to create
    (i.e., project) 2 different **1-mode** networks, one representing
    row x row members (e.g., student x student), with the cells
    representing some commonality calculated for the row pairs (using
    sums, averages, differences or correlation), and one representing
    the column x column entities (e.g., course x course).

# Useful Network Topologies

A network topology is the physical or logical arrangement of vertices
and edges within the network. These are useful when comparing networks
or when evaluating properties of networks. For example, a star
topology has one actor in the center with all other actors connected
to it, but not connected to each other. This topology is important
because it represents the most centralized network for a given network
size. These topologies typically used for computations of network
attributes, e.g., network centralization. These are some popular
topologies:

-   **Fully Connected:** every vertex is connected to every other
    vertex. The distance from a member to any other member is exactly
    one tie. A fully connected network has the maximum density
    possible for a network of a given size.

-   **Ring:** a vertex is connected to the next vertex, which in turn
    is connected to the next vertex, and so on, until the last vertex
    is connected to the first vertex, closing the loop.

-   **Star:** it contains one central vertex with all other vertex
    connected to it, but not connected to any other vertex. A star
    network of n vertices has one central vertex with a degree
    centrality of n-1 and the remaining vertices with a degree
    centrality of 1.

-   **Disconnected:** when there are no paths to connect some members
    to other members, the network is said to be disconnected. A
    connected network has one component, which is the actual network.
    The members have paths to all other members, although the network
    may not be fully connected. A disconnected network has multiple
    components. Components are disconnected from other components.

-   **Clustered:** all members are connected in triads. A member can
    be connected with many triads, but it must be connected to at
    least one triad. It is essentially a network with nothing but
    triangles.

-   **Hierarchical:** it is like a collection of more than one star
    networks and they all have directed ties. Every member has either
    no or many child vertices. The vertices without child vertices are
    the terminal vertices. Every vertex has only one parent vertex,
    except for the top-most vertex, which has no parent. The best way
    to visualize a hierarchical network is to think of it as an
    organizational chart, that is every one has subordinates, except
    the employees at the lowest level, and every one has a boss,
    except the top boss.

# Network Analysis Methods

There are two main types of network methods: quantitative and visual.
In addition, there are a number of transformations that can be applied
to network data to unleash the great power of the various methods.

-   **Quantitative:** this method relies on matrix manipulations and
    quantitative and statistical methods. Relationships in a network
    can be represented as binary or weighted inside a sociomatrix,
    which can then be inputed into the various quantitative methods
    available.

-   **Visual:** sociomatrices can be graphed and then analyzed using
    various visual methods. For example: vertices can be sized based
    on their centrality or other attribute; edges can be heavier or
    lighter depending on weighted edge attributes; communities and
    clusters can be colored and reported in informational visual
    graphs.

-   **Transformations:** these are not methods per se, but a clever
    transformation can lead into unique insights from the data. We
    will discuss many of these in later chapters. Some examples of
    useful transformations include:

    -   **Matricizing:** it involves converting non-network or
        non-matrix data into matrix data. A simple example is when you
        have a number of actors of different ages and having different
        incomes. You could use this data to construct a matrix that
        shows age or income differences between members. You can not
        only use differences, but you can also use other values like
        sums, averages, etc. Another example is to convert text to
        relations. For example, in the Network of Thrones example,
        analysts computed how many time every 2 actors in the novel
        appeared together in the text no more than 15 words appart,
        and used this count as a proximity value between them.

    -   **2-Mode to 1-Mode Projections:** for example, if you have a
        2-mode network containing students (rows) by courses
        (columns), with a 1 if the row student took the column course.
        You can then project this into a 1-mode student x student
        network, with cells containing how many times the row and
        column students took the same course.

    -   **Aggregate, dichotomize, group, etc.:** sky is the limit when
        it comes to transformations.

# Levels of Network Analysis

One of the strengths of network analysis is that there are many levels
of analysis offering a rich variety of perspectives, such as:

-   **Vertex (Node) Level:** collect or compute and analyze various
    node level and ego network attributes (centrality, isolates,
    etc.), such as:

    -   **Actor Prominence:** degree, betweenness or some other
        measure of centrality

    -   **Vertex Bridge:** when the only way to establish a path
        between two actors is through another actor, this actor is
        called a vertex bridge

    -   **Isolates:** vertices with no connections to others

    -   **Ego Network:** one actor's ties to neighboring actors
        (1-degree, i.e., 1 tie away, or n-degree, i.e., up to n ties
        away)

    -   **Structural Hole:** an actor with many ties to actors that
        are not tied to each other (i.e., stars)

-   **Edge (Tie) Level:** collect or compute and analyze relationship
    data (strong/weak ties, tie bridges, reciprocity, transitivity,
    etc.), such as:

    -   **Edge Bridge:** when the only way to establish a path between
        two vertices is through an edge between them, this edge is
        called an edge bridge -- the two nodes in an edge bridge are
        also vertex bridges

    -   **Path:** a sequence of adjacent vertices and edges connecting
        two nodes. Sometimes there is a single path between two nodes;
        other times there are multiple paths.

    -   **Distance:** the number of ties in the shortest path between
        2 vertices, also known as the Geodesic. Note that if the
        network is disconnected into more than one component, the
        distance between some members will be indeterminate (i.e., the
        distance between 2 members across unconnected components is
        infinite).

    -   **Reciprocity:** it applies to directed networks only. : if A
        --\> B and B --\> A, the tie is reciprocal. More generally,
        the tie between two actors can be either: none existent,
        directional or reciprocal.

-   **Sub-Group Level:** identify and analyze sub-groups, such as:

    -   **Cliques:** these are N vertices fully connected (3-clique,
        4-clique, etc.)

    -   **Sub-Graphs:** a focal part of a graph (e.g., a subgroup of
        members of the same gender or country)

    -   **Components:** fully connected sub-graphs, disconnected from
        other sub-graphs

    -   **Clusters:** groups of nodes with close proximity or
        similarity

    -   **Communities:** similar to clusters. More specifically, they
        are groups of actors with many ties within the community and
        sparse ties to members in other communities.

    -   **Triads (Transitivity):** 3-actor cliques A --\> B, A --\> C
        and B --\> C (i.e., forming a triad or triangle). Related to
        this, if A --\> B and B --\> C and A --\> C, means that the
        relationship between A and C is transitive and A, B and C form
        a triad cluster. Note: A network with high transitivity (many
        triads) has a high clustering coefficient and, consequently,
        it is low on structural holes (few stars)

-   **Network Level:** compute and analyze key network structural
    parameters, such as:

    -   **Network Size:** Number of vertices. Some social networks are
        small and specific (e.g., 12-member team) and can be analyzed
        as a single network. Sometimes you have many teams, and you
        want to predict team outcomes (e.g., 50 small teams of average
        size of 5). Sometimes you have very large groups (e.g.,
        communities of practice, social media groups) with thousands
        of nodes; visualization and data collection are more
        difficult.

    -   **Network Density:** Number of ties (edges), or the proportion
        of ties relative to the maximum number of possible ties for a
        network of a given size. Some networks are sparse (few
        ties/lots of weak ties) and some are dense (many ties/lots of
        strong ties).

    -   **Diameter:** is largest distance (geodesic) between actors in
        the network

    -   **Other:** Centralization, Isolation, Clustering, Hierarchy,
        Structure, etc.

-   **Multiplex Networks:** extract simplex network layers from a
    multiplex network for each relationship of interest and analyze
    their relationships or associations separately (e.g., correlation,
    regression):

    -   **Vertex Predictors** (gender, centrality) --\> **Vertex
        Outcomes** (success)

    -   **Edge Relationships** (supervisor, advice) --\> **Edge
        Outcomes** (friendship)

    -   **Mixed** (Vertex --\> Edge; Edge --\> Vertex)
